---
title: "Lici_Díaz.Rmd"
author: "Lici Díaz"
date: "5/12/2022"
output:
  pdf_document: default
  html_document: default
---

```{r}
#install.packages("reshape2") 
#install.packages("reshape")
#install.packages("nycflights13")
#install.packages("tableone")
#install.packages("rvest")
install.packages("yarrr")
install.packages("aqp")
install.packages("soilDB")
install.packages("package")

library(aqp)
library(soilDB)
library(tidyr)
library(MASS) 
library(reshape2) 
library(reshape)
library(Hmisc)
library(psych)
library(ggpubr)
library(readr)
library(tibble)
library(dplyr) 
library(magrittr)
library(lubridate)
library(nycflights13)
library(tableone)
library(survival)
library("yarrr")
library("ggplot2")
library("scales")
```

*Dataframe from csv files:*

```{r example_lab_data, synthetic_data}
example_lab_data <- read.table('example_lab_data.csv', sep=',', header=TRUE)
head(example_lab_data)
synthetic_data <- read.table('synthetic_data.csv', sep=',', header=TRUE)
head(synthetic_data)
```

*Remove any extraneous rows or columns:*

*Ensure that columns with numeric variables don’t have characters*
#Count number of nulls per column

```{r}
nullsum_example_lab_data <- sapply(example_lab_data, function(x) sum(is.na(x))) 
nullsum_synthetic_data <- sapply(synthetic_data, function(x) sum(is.na(x)))

```
#drop columns with NA > 300 observations:

```{r}
synthetic_data <- synthetic_data[,!names(synthetic_data) %in% c("diagnosis_icd10","supplemental_diagnoses", "bsa", "distal_perfusion",
 "trauma", "infective_endocarditis",  "severity_score", "insurance",   "il_6_blockers", "antivirals","antimalarials", "number_pt_visits_total_hosp_stay","or_cannulation", "steroid_name", "complication_mechanical", "mcs12", "qal_ys", "pcs12", "monoclonal_ab_treatment","total_charges","or_cannulation", "complication_patient_limb", "complication_pulmonary", "complication_renal", "complication_hemorrhage", "complication_mechanical", "complication_neurological", "complication_cardiovascular", "complication_metabolic", "cpc_score")]
```


*Review patient IDs and address repeated patients:*


*For synthetic_data*

```{r}
repeated_patients_syntetic_data <- data.frame(table(synthetic_data$patient_id))
repeated_patients_syntetic_data <- repeated_patients_syntetic_data[repeated_patients_syntetic_data$Freq > 1,]
```

*tells you which patient_id occurred more than once:*

```{r}
synthetic_data[synthetic_data$patient_id %in% repeated_patients_syntetic_data$Var1[repeated_patients_syntetic_data$Freq > 1],]
```

*drop duplicate patient_id, use only the first (oldest) visit:*

#For synthetic_data, Sort dataframe by patient_id and admission_date:

```{r}
synthetic_data <- synthetic_data[order(synthetic_data$patient_id,synthetic_data$admission_date ),]
```

#Drop duplicates dataframe by patient_id:

```{r}
synthetic_data <- synthetic_data[!duplicated(synthetic_data$patient_id), ]
```

*Ensure that columns with numeric variables don’t have characters*

#Fill empty values with NA:

```{r}
example_lab_data[example_lab_data == ''] <- NA
synthetic_data[synthetic_data == ''] <- NA
example_lab_data[example_lab_data == 'UTC'] <- NA
synthetic_data[synthetic_data == 'UTC'] <- NA
example_lab_data[example_lab_data == 'NDA'] <- NA
synthetic_data[synthetic_data == 'NDA'] <- NA
example_lab_data[example_lab_data == 'N/A'] <- NA
synthetic_data[synthetic_data == 'N/A'] <- NA
```

#find character on numeric columns:

```{r}
synthetic_data[synthetic_data == 'Not Available' ] <- NA
synthetic_data[synthetic_data == 'no info from OSH' ] <- NA
synthetic_data[synthetic_data == 'No data avail' ] <- NA
synthetic_data[synthetic_data == 'Not Measured, pt coded' ] <- NA
synthetic_data[synthetic_data == 'not measured' ] <- NA

synthetic_data$ph <- parse_number(synthetic_data$ph)
synthetic_data
```

#Some variables will have more than one value in the same cell, separated by a comma. 

*Use average for mechanical_vent_days in days not hours*

#drop first row (just NA values)

```{r}
synthetic_data = synthetic_data[-1,]
```
*Evaluate if categorical variables have any category with a low frequency:*

#Uses frecuency tables:

#standarizes sex colum:

```{r}
synthetic_data$sex[synthetic_data$sex == 'Female'] <- "F"
synthetic_data$sex[synthetic_data$sex == 'Male'] <- "M"


table(synthetic_data$sex)


table(synthetic_data$race)

table(synthetic_data$diagnosis)

table(synthetic_data$reintubation)

table(synthetic_data$trached)


table(synthetic_data$systemic_anticoagulation_type)

table(synthetic_data$acute_kidney_injury)

table(synthetic_data$discharge_location)

synthetic_data$steroids[synthetic_data$steroids == 'Unk'] <- "unk"
synthetic_data$steroids[synthetic_data$steroids == 'unk'] <- NA
table(synthetic_data$steroids)


synthetic_data$infection[synthetic_data$infection == 'R/V/SARS CoV19'] <- 'SARS CoV19'
synthetic_data$infection[synthetic_data$infection == 'R/V/COVID-19'] <- 'SARS CoV19'
synthetic_data$infection[synthetic_data$infection == 'Hx of COVID19 - 1/2022'] <- 'SARS CoV19'
synthetic_data$infection[synthetic_data$infection == 'P/R/V/ Covid-19, P/R/GM+/ Strep sp, P/B/GM+/ Staph captis'] <- 'SARS CoV19'


synthetic_data$infection[synthetic_data$infection != 'SARS CoV19'] <- 'Other'

table(synthetic_data$infection)

table(synthetic_data$support_type)

table(synthetic_data$transfer)

table(synthetic_data$covid)

table(synthetic_data$pregnant)

table(synthetic_data$mechanical_vent_days)
```

#for example_lab_data:

```{r}
table(example_lab_data$lab_exam) 
```

*Make sure that time variables are consistently coded:*
#convert char column into posixct

```{r}
synthetic_data[['admission_date']] <- as.POSIXct(synthetic_data[['admission_date']],
                                   format = "%Y-%m-%d")
                                   

synthetic_data[['discharge_date']] <- as.POSIXct(synthetic_data[['discharge_date']],
                                   format = "%Y-%m-%d")

synthetic_data[['death_date']] <- as.POSIXct(synthetic_data[['death_date']],
                                   format = "%Y-%m-%d")
```

#for example_lab_data:

```{r}
example_lab_data[['date']] <- as.POSIXct(example_lab_data[['date']],
                                   format = "%Y-%m-%d")
```

*Evaluate categorical variables:*

#for that, first convert to factors:

```{r}
#example_lab_data$lab_exam <- as.factor(example_lab_data$lab_exam)     

synthetic_data$sex <- as.factor(synthetic_data$sex) 
synthetic_data$race  <- as.factor(synthetic_data$race) 
synthetic_data$diagnosis  <- as.factor(synthetic_data$diagnosis) 
synthetic_data$reintubation  <- as.factor(synthetic_data$reintubation) 
synthetic_data$trached  <- as.factor(synthetic_data$trached) 

synthetic_data$systemic_anticoagulation_type <- as.factor(synthetic_data$systemic_anticoagulation_type) 

synthetic_data$acute_kidney_injury <- as.factor(synthetic_data$acute_kidney_injury)
 
synthetic_data$discharge_location <- as.factor(synthetic_data$discharge_location)
 
synthetic_data$steroids <- as.factor(synthetic_data$steroids)
synthetic_data$infection <- as.factor(synthetic_data$infection)
synthetic_data$support_type <- as.factor(synthetic_data$support_type)
synthetic_data$transfer <- as.factor(synthetic_data$transfer)
synthetic_data$covid <- as.factor(synthetic_data$covid)
synthetic_data$pregnant <- as.factor(synthetic_data$pregnant)
synthetic_data$mechanical_vent_days <- as.factor(synthetic_data$mechanical_vent_days)
```

*Evaluate the distribution of numeric variables:*

*We can evaluate normality according to swek and kurtosis:*
#statistical summary number variables:

```{r}
describe(synthetic_data$weight_kg)
describe(synthetic_data$height_cm)
describe(synthetic_data$bmi)
describe(synthetic_data$ph)
describe(synthetic_data$co2)
describe(synthetic_data$o2)
describe(synthetic_data$lactate_peak)
describe(synthetic_data$creatinine_peak)
describe(synthetic_data$total_bilirubin_peak)
describe(synthetic_data$hospital_los)
describe(synthetic_data$days_to_discharge)
```

*normal graph*

#As all the points fall approximately along this reference line, we can assume normality.

``` {r synthetic_data}
ggqqplot(synthetic_data$weight_kg)
ggqqplot(synthetic_data$height_cm)
ggqqplot(synthetic_data$bmi)
ggqqplot(synthetic_data$ph)
ggqqplot(synthetic_data$co2)
ggqqplot(synthetic_data$o2)
ggqqplot(synthetic_data$lactate_peak)
ggqqplot(synthetic_data$creatinine_peak)
ggqqplot(synthetic_data$total_bilirubin_peak)
ggqqplot(synthetic_data$hospital_los)
ggqqplot(synthetic_data$days_to_discharge)
```

*not normal numeric variables:*

synthetic_data$bmi
synthetic_data$o2
synthetic_data$lactate_peak
synthetic_data$total_bilirubin_peak
synthetic_data$hospital_los
synthetic_data$days_to_discharge

*If a numeric variable does not have a normal distribution, you may log transform it and re-evaluate the distribuion:*

#log transformation:

```{r}
synthetic_data$bmi_log = log(synthetic_data$bmi)
```

#graph to evaluate normality

```{r synthetic_data log}
ggqqplot(synthetic_data$bmi_log)


synthetic_data$o2_log = log(synthetic_data$o2)
ggqqplot(synthetic_data$o2_log)

synthetic_data$lactate_peak_log = log(synthetic_data$lactate_peak)
ggqqplot(synthetic_data$lactate_peak_log)

synthetic_data$total_bilirubin_peak_log = log(synthetic_data$total_bilirubin_peak)
ggqqplot(synthetic_data$total_bilirubin_peak_log)

synthetic_data$hospital_los_log = log(synthetic_data$hospital_los)
ggqqplot(synthetic_data$hospital_los_log)

synthetic_data$days_to_discharge_log = log(synthetic_data$days_to_discharge)
ggqqplot(synthetic_data$days_to_discharge_log)
```
*Bonus challange: Convert dataframe to tibble:*

```{r}
synthetic_data <- as.tbl(synthetic_data)
example_lab_data <- as.tbl(example_lab_data)
```

*To link the example_lab_data with synthetic_data i decide that is not well done calculate the mean of the values, because there are taken on too many different times, and some values are not even near to each other, so I decide to link it by the year and month and patient_id*


#Create a column "month" and "year" for synthetic data:
```{r}

synthetic_data$month_discharge = month(as.POSIXlt(synthetic_data$discharge_date, format="%d/%m/%Y"))

synthetic_data$month_admission = month(as.POSIXlt(synthetic_data$admission_date, format="%d/%m/%Y"))

synthetic_data$year_admission = year(as.POSIXlt(synthetic_data$admission_date, format="%d/%m/%Y"))

synthetic_data$year_discharge = year(as.POSIXlt(synthetic_data$discharge_date, format="%d/%m/%Y"))

```
#Create a column "year" and "month" for example_lab_data:

```{r}
example_lab_data$month = month(as.POSIXlt(example_lab_data$date, format="%d/%m/%Y"))
example_lab_data$year = year(as.POSIXlt(example_lab_data$date, format="%d/%m/%Y"))

```
#cchange format for example_lab_data from long to wide:
```{r}
example_lab_data_wide <- pivot_wider(
  example_lab_data,
  id_cols = NULL,
  id_expand = FALSE,
  names_from = lab_exam,
  values_from = result)
print(example_lab_data_wide)
```

*Left join synthetic_data with example_lab_data_wide:*

#This join able to concate all the data, for extract coherent data for the time that the patient was on the hospital, could be based on admission_date or discharge_date:

```{r}
linkdataset <- merge(x = synthetic_data, y = example_lab_data_wide, by = "patient_id")
```
#to obtaind coherent lab_exam values, gonna choose those lab exam who was made the year when the patient was on the hospital and drop the others. 

```{r}
linkdataset_year <- subset(linkdataset, linkdataset$year.y == linkdataset$year_admission)
```

#And with linkdataset_year, gonna choose those patient_id where the month of the lab exam was taken between the months of the patients on the hospital, it means between the month of admission or discharche, that because is is not coherent take information about the lab exams about the patient where they was not on the hospital, conditions of the patient could change between months and years, some values are too far from each other to make an average. This is an small dataset, but this information is enough trusted to realize if those lab exams results have relationship with the output variable of the model, on other way, we can take the large dataset without this lab exams, but after check correlation between the variables:

```{r}
linkdataset_month <- subset(linkdataset_year, linkdataset$month >= linkdataset$month_admission & linkdataset$month <= linkdataset$month_discharge)

linkdataset_month <- subset(linkdataset_year, linkdataset_year$month >= linkdataset_year$month_admission & linkdataset_year$month <= linkdataset_year$month_discharge)

linkdataset_month

```
**Exploratory analysis**

*Summary table stratified by covid diagnosis:*

#first have to separate variable between numeric and categorical:

# first prepare the function variables:

```{r}

#all variables except those who aren't part of the satistical analysis:
myVars <- colnames(linkdataset_month[,!names(linkdataset_month) %in% c("covid","patient_id", "year.y", "month_discharge","month_admission", "year_admission", "year_discharge", "month", "year.x" )])

#select categorical variables:
catVars <- c("sex", "race", "diagnosis", "reintubation", "trached", "systemic_anticoagulation_type", "acute_kidney_injury", "discharge_location", "steroids", "infection", "support_type", "transfer", "pregnant", "mechanical_vent_days")

#wasn't necessary use "factorVars" argument, because i already did the factor trasnformation for the cat variables on line 215 -237, there isn't also not normal variables, thaks to the log transformation:

#Summary table:
tab <- CreateTableOne(vars = myVars, strata = "covid", data = linkdataset_month)

print(tab, formatOptions = list(big.mark = ","))

summary(tab)

```

*Create a pirate plot as well as a scatterplot with a smooth spline using any
variables of your choice:*


```{r}
theme_economist(
  base_size = 10,
  base_family = "sans",
  horizontal = TRUE,
  dkpanel = FALSE
)

theme_economist_white(
  base_size = 11,
  base_family = "sans",
  gray_bg = TRUE,
  horizontal = TRUE
)
```
```{r}
#pireplot

pirateplot(formula = icu_los ~ infection, 
           data = linkdataset_month,
           pal = "pony", 
           main = "Covid vs CO2") 



#interpretation: 

#CO2 levels on patients diagnosticated with COVID-19 are slowly higher that those who was negative for COVID-19. 

```

```{r}
#scatterplot:

x <- linkdataset_month$icu_los
y <- linkdataset_month$co2


# Plot with main and axis titles
# Change point shape (pch = 19) and remove frame.
plot(x, y, main = "icu_los vs co2",
     xlab = "icu_los", ylab = "co2",
     pch = 19, frame = FALSE,
     #xlim = c(20, 60),
     ylim = c(0, 150), theme = "The economist")

#What we can analyze from this graph that when co2 is higer, more days on icu could be the patient, but is not too clear the relationship.

```
# chi square test to evaluate correlation between catogorical variables of synthetic data:
```{r}

#H0:the two variables are independent, Ha: Two variables are dependent, if p-value < 0.05 so H0 has to be rejected:

table(synthetic_data$sex, synthetic_data$covid)
chisq.test(synthetic_data$sex, synthetic_data$covid, correct=FALSE)


#We have a chi-squared value of 0.06604. Since we get a p-Value higer than the significance level of 0.05, we can't reject the null hypothesis and conclude that the two variables are in fact independent.

table(synthetic_data$race, synthetic_data$covid)
chisq.test(synthetic_data$race, synthetic_data$covid, correct=FALSE)

table(synthetic_data$diagnosis, synthetic_data$covid)
chisq.test(synthetic_data$diagnosis, synthetic_data$covid, correct=FALSE)

table(synthetic_data$reintubation, synthetic_data$covid)
chisq.test(synthetic_data$reintubation, synthetic_data$covid, correct=FALSE)

table(synthetic_data$trached, synthetic_data$covid)
chisq.test(synthetic_data$trached, synthetic_data$covid, correct=FALSE)

table(synthetic_data$mechanical_vent_days, synthetic_data$covid)
chisq.test(synthetic_data$mechanical_vent_days, synthetic_data$covid, correct=FALSE)

table(synthetic_data$systemic_anticoagulation_type, synthetic_data$covid)
chisq.test(synthetic_data$systemic_anticoagulation_type, synthetic_data$covid, correct=FALSE)

table(synthetic_data$acute_kidney_injury, synthetic_data$covid)
chisq.test(synthetic_data$acute_kidney_injury, synthetic_data$covid, correct=FALSE)

table(synthetic_data$discharge_location, synthetic_data$covid)
chisq.test(synthetic_data$discharge_location, synthetic_data$covid, correct=FALSE)

table(synthetic_data$steroids, synthetic_data$covid)
chisq.test(synthetic_data$steroids, synthetic_data$covid, correct=FALSE)

table(synthetic_data$infection, synthetic_data$covid)
chisq.test(synthetic_data$infection, synthetic_data$covid, correct=FALSE)

table(synthetic_data$support_type, synthetic_data$covid)
chisq.test(synthetic_data$support_type, synthetic_data$covid, correct=FALSE)

table(synthetic_data$transfer, synthetic_data$covid)
chisq.test(synthetic_data$transfer, synthetic_data$covid, correct=FALSE)

```
*conclude: just steroids, and mechanical_vent_days are related with covid diagnosis.*

*To analyze numerical variables, we can use a scatterplot matrix, to evaluate linear correlation between some variables:*
```{r}
#data <- (linkdataset_month[,!names(linkdataset_month) %in% #c("covid","patient_id", "year.y", "month_discharge","month_admission", #"year_admission", "year_discharge", "month", "year.x" )])
  
  #c("weight_kg", "height_cm", "ph", "co2", "icu_los", "bmi_log", "o2_log")

plot(data , pch=20 , cex=1.5 , col="#69b3a2")

data1 <- (linkdataset_month[,!names(linkdataset_month) %in% c("covid","patient_id", "year.y", "month_discharge","month_admission", "year_admission", "year_discharge", "month", "year.x", "weight_kg", "height_cm", "ph", "co2", "icu_los", "bmi_log", "o2_log", "sex", "race", "diagnosis", "reintubation", "trached", "systemic_anticoagulation_type", "acute_kidney_injury", "discharge_location", "steroids", "infection", "support_type", "transfer", "pregnant", "mechanical_vent_days", "proning", "total_bilirubin_peak","o2", "days_to_discharge","admission_date", "discharge_date", "death_date", "date", "Hematocrit", "Cholesterol", "CRP", "Platelets", "Glucose", "Lymphocytes", "bmi", "bmi_log", "hospital_los_log", "lactate_peak")])

plot(data1 , pch=20 , cex=1.5 , col="#69b3a2")

```
*Numeric variables could be related:*
 For example creatinine peak and hospital los could be related because, less values on creatinine less hospital days.

*This  is helpful to build the GML model.


