---
title: "Lici_Díaz.Rmd"
author: "Lici Díaz"
date: "5/12/2022"
output:
  pdf_document: default
  html_document: default
---

```{r}
install.packages("reshape2") 
install.packages("reshape")
install.packages("nycflights13")
library(tidyr)
library(MASS) 
library(reshape2) 
library(reshape)
library(Hmisc)
library(psych)
library(ggpubr)
library(readr)
library(tibble)
library(dplyr) 
library(magrittr)
library(lubridate)
library(nycflights13)
```

*Dataframe from csv files:*

```{r example_lab_data, synthetic_data}
example_lab_data <- read.table('example_lab_data.csv', sep=',', header=TRUE)
head(example_lab_data)
synthetic_data <- read.table('synthetic_data.csv', sep=',', header=TRUE)
head(synthetic_data)
```

*Remove any extraneous rows or columns:*

*Ensure that columns with numeric variables don’t have characters*
#Count number of nulls per column

```{r}
nullsum_example_lab_data <- sapply(example_lab_data, function(x) sum(is.na(x))) 
nullsum_synthetic_data <- sapply(synthetic_data, function(x) sum(is.na(x)))

```
#drop columns with NA > 300 observations:

```{r}
synthetic_data <- synthetic_data[,!names(synthetic_data) %in% c("diagnosis_icd10","supplemental_diagnoses", "bsa", "distal_perfusion",
 "trauma", "infective_endocarditis",  "severity_score", "icu_los","insurance", "proning",  "il_6_blockers", "antivirals","antimalarials", "number_pt_visits_total_hosp_stay","or_cannulation", "steroid_name", "complication_mechanical", "mcs12", "qal_ys", "pcs12", "monoclonal_ab_treatment","total_charges","or_cannulation", "complication_patient_limb", "complication_pulmonary", "complication_renal", "complication_hemorrhage", "complication_mechanical", "complication_neurological", "complication_cardiovascular", "complication_metabolic", "cpc_score")]
```


*Review patient IDs and address repeated patients:*


*For synthetic_data*

```{r}
repeated_patients_syntetic_data <- data.frame(table(synthetic_data$patient_id))
repeated_patients_syntetic_data <- repeated_patients_syntetic_data[repeated_patients_syntetic_data$Freq > 1,]
```

*tells you which patient_id occurred more than once:*

```{r}
synthetic_data[synthetic_data$patient_id %in% repeated_patients_syntetic_data$Var1[repeated_patients_syntetic_data$Freq > 1],]
```

*drop duplicate patient_id, use only the first (oldest) visit:*

#For synthetic_data, Sort dataframe by patient_id and admission_date:

```{r}
synthetic_data <- synthetic_data[order(synthetic_data$patient_id,synthetic_data$admission_date ),]
```

#Drop duplicates dataframe by patient_id:

```{r}
synthetic_data <- synthetic_data[!duplicated(synthetic_data$patient_id), ]
```

*Ensure that columns with numeric variables don’t have characters*

#Fill empty values with NA:

```{r}
example_lab_data[example_lab_data == ''] <- NA
synthetic_data[synthetic_data == ''] <- NA
example_lab_data[example_lab_data == 'UTC'] <- NA
synthetic_data[synthetic_data == 'UTC'] <- NA
example_lab_data[example_lab_data == 'NDA'] <- NA
synthetic_data[synthetic_data == 'NDA'] <- NA
example_lab_data[example_lab_data == 'N/A'] <- NA
synthetic_data[synthetic_data == 'N/A'] <- NA
```

#find character on numeric columns:

```{r}
synthetic_data[synthetic_data == 'Not Available' ] <- NA
synthetic_data[synthetic_data == 'no info from OSH' ] <- NA
synthetic_data[synthetic_data == 'No data avail' ] <- NA
synthetic_data[synthetic_data == 'Not Measured, pt coded' ] <- NA
synthetic_data[synthetic_data == 'not measured' ] <- NA

synthetic_data$ph <- parse_number(synthetic_data$ph)
synthetic_data
```

#Some variables will have more than one value in the same cell, separated by a comma. 

*Use average for mechanical_vent_days in days not hours*

#drop first row (just NA values)

```{r}
synthetic_data = synthetic_data[-1,]
```
*Evaluate if categorical variables have any category with a low frequency:*

#Uses frecuency tables:

#standarizes sex colum:

```{r}
synthetic_data$sex[synthetic_data$sex == 'Female'] <- "F"
synthetic_data$sex[synthetic_data$sex == 'Male'] <- "M"


table(synthetic_data$sex)


table(synthetic_data$race)

table(synthetic_data$diagnosis)

table(synthetic_data$reintubation)

table(synthetic_data$trached)


table(synthetic_data$systemic_anticoagulation_type)

table(synthetic_data$acute_kidney_injury)

table(synthetic_data$discharge_location)

synthetic_data$steroids[synthetic_data$steroids == 'Unk'] <- "unk"
synthetic_data$steroids[synthetic_data$steroids == 'unk'] <- NA
table(synthetic_data$steroids)


synthetic_data$infection[synthetic_data$infection == 'R/V/SARS CoV19'] <- 'SARS CoV19'
synthetic_data$infection[synthetic_data$infection == 'R/V/COVID-19'] <- 'SARS CoV19'
synthetic_data$infection[synthetic_data$infection == 'Hx of COVID19 - 1/2022'] <- 'SARS CoV19'
synthetic_data$infection[synthetic_data$infection == 'P/R/V/ Covid-19, P/R/GM+/ Strep sp, P/B/GM+/ Staph captis'] <- 'SARS CoV19'


synthetic_data$infection[synthetic_data$infection != 'SARS CoV19'] <- 'Other'

table(synthetic_data$infection)

table(synthetic_data$support_type)

table(synthetic_data$transfer)

table(synthetic_data$covid)

table(synthetic_data$pregnant)

table(synthetic_data$mechanical_vent_days)
```

#for example_lab_data:

```{r}
table(example_lab_data$lab_exam) 
```

*Make sure that time variables are consistently coded:*
#convert char column into posixct

```{r}
synthetic_data[['admission_date']] <- as.POSIXct(synthetic_data[['admission_date']],
                                   format = "%Y-%m-%d")
                                   

synthetic_data[['discharge_date']] <- as.POSIXct(synthetic_data[['discharge_date']],
                                   format = "%Y-%m-%d")

synthetic_data[['death_date']] <- as.POSIXct(synthetic_data[['death_date']],
                                   format = "%Y-%m-%d")
```

#for example_lab_data:

```{r}
example_lab_data[['date']] <- as.POSIXct(example_lab_data[['date']],
                                   format = "%Y-%m-%d")
```

*Evaluate categorical variables:*

#for that, first convert to factors:

```{r}
#example_lab_data$lab_exam <- as.factor(example_lab_data$lab_exam)     

synthetic_data$sex <- as.factor(synthetic_data$sex) 
synthetic_data$race  <- as.factor(synthetic_data$race) 
synthetic_data$diagnosis  <- as.factor(synthetic_data$diagnosis) 
synthetic_data$reintubation  <- as.factor(synthetic_data$reintubation) 
synthetic_data$trached  <- as.factor(synthetic_data$trached) 

synthetic_data$systemic_anticoagulation_type <- as.factor(synthetic_data$systemic_anticoagulation_type) 

synthetic_data$acute_kidney_injury <- as.factor(synthetic_data$acute_kidney_injury)
 
synthetic_data$discharge_location <- as.factor(synthetic_data$discharge_location)
 
synthetic_data$steroids <- as.factor(synthetic_data$steroids)
synthetic_data$infection <- as.factor(synthetic_data$infection)
synthetic_data$support_type <- as.factor(synthetic_data$support_type)
synthetic_data$transfer <- as.factor(synthetic_data$transfer)
synthetic_data$covid <- as.factor(synthetic_data$covid)
synthetic_data$pregnant <- as.factor(synthetic_data$pregnant)
synthetic_data$mechanical_vent_days <- as.factor(synthetic_data$mechanical_vent_days)
```

*Evaluate the distribution of numeric variables:*

*We can evaluate normality according to swek and kurtosis:*
#statistical summary number variables:

```{r}
describe(synthetic_data$weight_kg)
describe(synthetic_data$height_cm)
describe(synthetic_data$bmi)
describe(synthetic_data$ph)
describe(synthetic_data$co2)
describe(synthetic_data$o2)
describe(synthetic_data$lactate_peak)
describe(synthetic_data$creatinine_peak)
describe(synthetic_data$total_bilirubin_peak)
describe(synthetic_data$hospital_los)
describe(synthetic_data$days_to_discharge)
```

*normal graph*

#As all the points fall approximately along this reference line, we can assume normality.

``` {r synthetic_data}
ggqqplot(synthetic_data$weight_kg)
ggqqplot(synthetic_data$height_cm)
ggqqplot(synthetic_data$bmi)
ggqqplot(synthetic_data$ph)
ggqqplot(synthetic_data$co2)
ggqqplot(synthetic_data$o2)
ggqqplot(synthetic_data$lactate_peak)
ggqqplot(synthetic_data$creatinine_peak)
ggqqplot(synthetic_data$total_bilirubin_peak)
ggqqplot(synthetic_data$hospital_los)
ggqqplot(synthetic_data$days_to_discharge)
```

*not normal numeric variables:*

synthetic_data$bmi
synthetic_data$o2
synthetic_data$lactate_peak
synthetic_data$total_bilirubin_peak
synthetic_data$hospital_los
synthetic_data$days_to_discharge

*If a numeric variable does not have a normal distribution, you may log transform it and re-evaluate the distribuion:*

#log transformation:

```{r}
synthetic_data$bmi_log = log(synthetic_data$bmi)
```

#graph to evaluate normality

```{r synthetic_data log}
ggqqplot(synthetic_data$bmi_log)


synthetic_data$o2_log = log(synthetic_data$o2)
ggqqplot(synthetic_data$o2_log)

synthetic_data$lactate_peak_log = log(synthetic_data$lactate_peak)
ggqqplot(synthetic_data$lactate_peak_log)

synthetic_data$total_bilirubin_peak_log = log(synthetic_data$total_bilirubin_peak)
ggqqplot(synthetic_data$total_bilirubin_peak_log)

synthetic_data$hospital_los_log = log(synthetic_data$hospital_los)
ggqqplot(synthetic_data$hospital_los_log)

synthetic_data$days_to_discharge_log = log(synthetic_data$days_to_discharge)
ggqqplot(synthetic_data$days_to_discharge_log)
```
*Bonus challange: Convert dataframe to tibble:*

```{r}
synthetic_data <- as.tbl(synthetic_data)
example_lab_data <- as.tbl(example_lab_data)
```

*To link the example_lab_data with synthetic_data i decide that is not well done calculate the mean of the values, because there are taken on too many different times, and some values are not even near to each other, so I decide to link it by the year and month and patient_id*


#Create a column "month" and "year" for synthetic data:
```{r}

#synthetic_data <- synthetic_data %>%
#  add_column(month_admission = month(as.POSIXlt(synthetic_data$admission_date, #format="%d/%m/%Y")))

#synthetic_data <- synthetic_data[,!names(synthetic_data) %in% c("year_admission ","month_admission", "month_discharge","year_admission", "year_discharge")]


synthetic_data$month_discharge = month(as.POSIXlt(synthetic_data$discharge_date, format="%d/%m/%Y"))

synthetic_data$month_admission = month(as.POSIXlt(synthetic_data$admission_date, format="%d/%m/%Y"))

synthetic_data$year_admission = year(as.POSIXlt(synthetic_data$admission_date, format="%d/%m/%Y"))

synthetic_data$year_discharge = year(as.POSIXlt(synthetic_data$discharge_date, format="%d/%m/%Y"))

```
#Create a column "year" and "month" for example_lab_data:

```{r}
example_lab_data$month = month(as.POSIXlt(example_lab_data$date, format="%d/%m/%Y"))
example_lab_data$year = year(as.POSIXlt(example_lab_data$date, format="%d/%m/%Y"))

```
#cchange format for example_lab_data from long to wide:
```{r}
example_lab_data_wide <- pivot_wider(
  example_lab_data,
  id_cols = NULL,
  id_expand = FALSE,
  names_from = lab_exam,
  values_from = result)
print(example_lab_data_wide)
```

*Left join synthetic_data with example_lab_data_wide:*

#This join able to concate all the data, for extract coherent data for the time that the patient was on the hospital, could be based on admission_date or discharge_date:

```{r}
linkdataset <- merge(x = synthetic_data, y = example_lab_data_wide, by = "patient_id")
```
#to obtaind coherent lab_exam values, gonna choose those lab exam who was made when the patient was on the hospital and drop the others. 

```{r}
f <- subset(linkdataset, linkdataset$year.y == linkdataset$year_admission)

g <- subset(f, linkdataset$month >= linkdataset$month_admission & linkdataset$month <= linkdataset$month_discharge)

g


```
